msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-08 paper.tex
Update bottleneck section.

\subsection*{Finite sites mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and
GTR~\citep{tavare1986some} nucleotide models
and
the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}, and are run before every
release.

@@

Kingman's coalescent assumes that only two ancestral lineages can merge at
each merger event. Although this is generally a reasonable approximation there
are certain situations in which the underlying mathematical assumptions are
violated. For example in certain highly fecund organisms
\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
to produce numbers of offspring on the order of the population size and
therefore a few individuals may produce the bulk of the offspring in any given
generation\citep{hedgecock_94}. These population dynamics violate basic
assumptions of the Kingman coalescent, and are better modelled by
`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
two lineages can merge in a given event. Multiple-merger coalescent processes
have also been shown to be relevant for modeling the effects of selection on
gene genealogies \citep{Gillespie909,DS04}.

@@

\subsection*{Instantaneous bottlenecks}
A common approach to modelling the effect of demographic history on genealogies is to assume that effective population size ($N_e$) changes in discrete steps which define a series of epochs~\citep{griffiths1994sampling, marth2004allele, keightley2007joint, li2011inference}. In this setting of piece-wise constant $N_e$, capturing a population bottleneck requires three epochs: $N_e$ is reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If bottlenecks are short both on the timescale of coalescence and mutations, one may expect little information about the duration of a bottleneck ($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$) and generate a sudden burst of coalescence events (a multiple merger event) in the genealogy. The strength of the bottleneck B can be thought of as an (imaginary) time period during which coalescence events are collapsed, i.e. there is no growth in genealogical branches during B and the probability that a single pair of lineages entering the bottleneck coalesce during the bottleneck is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is attractive and both analytic results and empirical inference~\citep{griffiths1994sampling, galtier2000detecting, bunnefeld2015inferring} have been developed under this model, it has not been straightforward to simulate data under instantaneous bottleneck histories with \ms. Instantaneous bottlenecks are implemented as a demographic event in \msprime.

>>>>

\subsection*{Finite sites mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

Kingman's coalescent assumes that only two ancestral lineages can merge at
each merger event. Although this is generally a reasonable approximation there
are certain situations in which the underlying mathematical assumptions are
violated. For example in certain highly fecund organisms
\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
to produce numbers of offspring on the order of the population size and
therefore a few individuals may produce the bulk of the offspring in any given
generation~\citep{hedgecock_94}. These population dynamics violate basic
assumptions of the Kingman coalescent, and are better modelled by
`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
two lineages can merge in a given event. Multiple-merger coalescent processes
have also been shown to be relevant for modeling the effects of selection on
gene genealogies \citep{Gillespie909,DS04}.

@@

\subsection*{Instantaneous bottlenecks}

A common approach to modelling the effect of demographic history on genealogies
is to assume that effective population size ($N_e$) changes in discrete steps
which define a series of epochs~\citep{griffiths1994sampling, marth2004allele,
keightley2007joint,li2011inference}. In this setting of piecewise constant
$N_e$, capturing a population bottleneck requires three epochs: $N_e$ is
reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and
recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If
bottlenecks are short both on the timescale of coalescence and mutations,
there may be little information about the duration of a bottleneck
($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to
assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$)
and generate a sudden burst of coalescence events (a multiple merger event) in
the genealogy. The strength of the bottleneck $B$ can be thought of as an
(imaginary) time period during which coalescence events are collapsed,
i.e.\ there is no growth in genealogical branches during $B$ and the probability that a
single pair of lineages entering the bottleneck coalesce during the bottleneck
is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is
attractive and both analytic results and empirical
inference~\citep{griffiths1994sampling, galtier2000detecting,
bunnefeld2015inferring} have been developed under this model, there has
been no software available to simulate data under instantaneous
bottleneck histories.

We have implemented instantaneous bottlenecks in \msprime~1.0
using a variant of Hudson's linear time single-locus coalescent
algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
by adding events to the Demography object (see the Demography section)
and can be used in combination with any other demographic modelling
features. We have validated the results of these simulations by comparing
against analytical expectations for coalescence times and the
site frequency spectrum~\citep{bunnefeld2015inferring}.
